An efficient composition of bidirectional programs by tupling and lazy updates

### 2019-11-27 paper/4.xpg.tex
fix indent + scaled tick

\section{Tupling and lazy updates: xpg}
\subsection{Tupling: pg}
s
Another solution for saving intermediate states is tupling. If $put$ and $get$ are evaluated simultaneously, it is potential to reduce the number of recomputed $get$s. The following function, $pg$, accepts the pair of a source and a view as the input to produce a new pair that contains the actual result of the corresponding minBiGUL program.

@@

Now, let's see how we construct $pg$ recursively.

$\tab \pg{Skip \ h}{s}{v} \overset{1}{=} (\sif{h \ s = v}{s}{\text{fail}}, h \ s) \\
    \qtab \overset{2}{=} \sif{h \ s = v}{(s, h \ s)}{\text{fail}} \\
    \qtab \overset{3}{=} \sif{h \ s = v}{(s, v)}{\text{fail}}$\\
The first equality is simply based on the definitions of $pg$, $\putbxinline{Skip \ h}$ and $\getbxinline{Skip \ h}$. The second one tuples two results of $put$ and $get$ in the body of the if-expression. This is a trick since in some cases, the result of $pg$ may be fail although there is no fail when evaluating $\getbxinline{Skip \ h}$. The last equality is relatively oblivious.

$\tab \pg{Replace}{s}{v} = (v, s)$\\
$\tab \pg{\product{bx_1}{bx_2}}{(s_1,s_2)}{(v_1,v_2)}\\
    \qtab \overset{1}{=} ((\putbx{bx_1}{s_1}{v_1}),(\putbx{bx_2}{s_2}{v_2}), (\getbx{bx_1}{s_1}), (\getbx{bx_2}{s_2})) \\
    \qtab \overset{2}{=} (s_1, v_1) \Leftarrow \pg{bx_1}{s_1}{v_1}; \\
        \qtab \tab \ \, (s_2, v_2) \Leftarrow \pg{bx_2}{s_2}{v_2}; \\
        \qtabs{2} ((s_1,s_2), (v_1,v_2))$\\
$\tab \pg{\rearrs{f_1}{f_2}{bx}}{s}{v} \overset{1}{=} (f_2 \ (\putbx{bx}{(f_1 \ s)}{v}), \getbx{bx}{(f_1 \ s)}) \\
    \qtab \overset{2}{=} (s, v) \Leftarrow \pg{bx}{f_1 \ s}{v};\\
        \qtabs{2} (f_2 \ s, v)$\\
$\tab \pg{\rearrv{g_1}{g_2}{bx}}{s}{v} \overset{1}{=} (\putbx{bx}{s}{(g_1 \ v)}, g_2 \ (\getbx{bx}{s})) \\
    \qtab \overset{2}{=} (s, v) \Leftarrow \pg{bx}{s}{g_1 \ v}; \\
        \qtabs{2} (s, g_2 \ v)$\\
Constructions of $pg$ for the replacement, the product and the source/view rearrangements are very clear when just paring $put$ and $get$ respectively, then doing basic changes.

$\tab \pg{\casebx{cond_{sv}}{cond_{s}}{bx_1}{bx_2}}{s}{v} \\
\tabs{2} \overset{1}{=} (\text{if} \ {cond_{sv} \ s \ v} \qtabs{3} \tab \text{if} \ {cond_{s} \ s} \\
\tab \qtab \text{then} \ s' \Leftarrow \putbx{bx_1}{s}{v} \qtab \tab \, \text{then} \ v' \Leftarrow \getbx{bx_1}{s} \\
\tab \qtab \text{else} \ s' \Leftarrow \putbx{bx_2}{s}{v} \qtab \tab \ \, \text{else} \ v' \Leftarrow \getbx{bx_2}{s}\\
\tab \qtab \text{fi} \ cond_{s} \ s'; \text{ return } s' \qtab , \tab \ \ \text{fi} \ {cond_{sv} \ s \ v'}; \text{ return } v')\\
\tabs{2} \overset{2}{=} \text{if} \ {cond_{sv} \ s \ v} \ \&\& \ {cond_{s} \ s}\\
\tab \qtab \text{then} \ (s', v') \Leftarrow \pg{bx_1}{s}{v}\\
\tab \qtab \text{else} \ (s', v') \Leftarrow \pg{bx_2}{s}{v}\\
\tab \qtab \text{fi} \ cond_{s} \ s' \ \&\& \ cond_{sv} \ s \ v'; \text{ return } (s',v')$

@@

$\tab \pg{bx_1 \circ bx_2}{s}{v} \\
    \tabs{2} \overset{1}{=} (\putbx{bx_1}{s}{(\putbx{bx_2}{(\getbx{bx_1}{s})}{v})}, \getbx{bx_2}{(\getbx{bx_1}{s})}) \\
    \tabs{2} \overset{2}{=} v_1 \Leftarrow \getbx{bx_1}{s}; \qtabs{3} \ \, \overset{3}{=} (s_1, v_1) \Leftarrow \pg{bx_1}{s}{dummy};\\
        \tab \qtab (s_2, v_2) \Leftarrow \pg{bx_2}{v_1}{v}; \qtabs{2} (s_2, v_2) \Leftarrow \pg{bx_2}{v_1}{v};\\
        \tab \qtab (s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2}; \qtabs{2} (s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2};\\
            \qtab \tabs{2} (s_3, v_2) \qtabs{6} (s_3, v_2)\\
    \tabs{2} \overset{4}{=} (s_1, v_1) \Leftarrow \pg{bx_1}{s}{construct\_dummy \ s}; \\
        \tab \qtab (s_2, v_2) \Leftarrow \pg{bx_2}{v_1}{v}; \\
        \tab \qtab (s_3, v_3) \Leftarrow \pg{bx_1}{s_1}{s_2}; \\
            \qtab \tabs{2} (s_3, v_2)$

            The construction of $\pginline{bx_1 \circ bx_2}$ can be considered as the soul of the $pg$ function. The first two equalities comes from mentioned definitions and some basic transformations. The third one rewrites $v_1 \Leftarrow \getbx{bx_1}{s}$ into $(s_1, v_1) \Leftarrow \pg{bx_1}{s}{dummy}$. It is possible if we consider $\getbx{bx_1}{s}$ as the second element of $\pg{bx_1}{s}{dummy}$ where $dummy$ is a special value that makes the $put \, [\![bx_1]\!]$ valid.We can construct a $dummy$ from the source by \textsc{GetPut} law in some cases or require a valid view from programmers.        
            The last equality changes $dummy$ by an application $construct\_dummy \ s$, and also substitutes $s$ in $(s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2}$ by $s_1$ which equals to $\putbx{bx_1}{s}{(construct\_dummy \ s)}$.
            Because there is no real view here, we need to construct a dummy from the source. 
            The replacement of $s$ and $s_1$ is possible because of the \textsc{PutPut} law. Then, $v_3$ in the result pair $(s_3,v_3)$ equals $dummy$. So we can realize that there is no loss information when computing a composition.

@@

$\cpg{Skip \ h}{ks}{kv}{s}{v} = \text{ if } h \ s = v \text{ then } (ks, kv, s, v) \text{ else fail}$

$\cpg{Replace}{ks}{kv}{s}{v} = (kv, ks, v, s)$

$\cpg{\product{bx_1}{bx_2}}{ks}{kv}{s}{v} =\\
    \tab (ks_1, kv_1, s_1, v_1) \Leftarrow \cpg{bx_1}{fst \circ ks}{fst \circ kv}{fst \ s}{fst \ v};\\
    \tab (ks_2, kv_2, s_2, v_2) \Leftarrow \cpg{bx_2}{snd \circ ks}{snd \circ kv}{snd \ s}{snd \ v};\\
    \qtab (con \ ks_1 \ ks_2, con \ kv_1 \ kv_2, (s_1,s_2), (v_1,v_2))$

$\cpg{RearrS \ f_1 \ f_2 \ bx}{ks}{kv}{s}{v} =\\
    \tab (ks, kv, s, v) \Leftarrow \cpg{bx}{f_1 \circ ks}{kv}{f_1 \ s}{v};\\
    \qtab (f_2 \circ ks, kv, s, v)$

$\cpg{RearrV \ g_1 \ g_2 \ bx}{ks}{kv}{s}{v} =\\
    \tab (ks, kv, s, v) \Leftarrow \cpg{bx}{ks}{g_1 \circ kv}{s}{g_1 \ v};\\
    \qtab (ks, g_2 \circ kv, ks, g_2 \ v)$

$\cpg{Case \ cond_{sv} \ cond_{s} \ bx_1 \ bx_2}{ks}{kv}{s}{v} =\\
    \tab \text{if} \ cond_{sv} \ s \ v \ \&\& \ cond_{s} \ s\\
    \tab \text{then} \ (ks, kv, s', v') \Leftarrow \cpg{bx_1}{ks}{kv}{s}{v}\\
    \tab \text{else} \ (ks, kv, s', v') \Leftarrow \cpg{bx_2}{ks}{kv}{s}{v}\\
    \tab \text{fi} \ cond_{s} \ s' \ \&\& \ cond_{sv} \ s \ v'; \ \text{return} \ (ks, kv, s', v')$

$\cpg{bx_1 \circ bx_2}{ks}{kv}{s}{v} =\\
    \tab (ks_1, kv_1, s_1, v_1) \Leftarrow \cpg{bx_1}{ks}{id}{s}{construct\_dummy \ s};\\
    \tab (ks_2, kv_2, s_2, v_2) \Leftarrow \cpg{bx_2}{kv_1}{kv}{v_1}{v};\\
        \qtab (ks_1 \circ ks_2, kv_2,  ks_1 \ s_2, v_2)$
\end{definition}

Apart from the last construction, the others are quite similar to the corresponding ones of $pg$, but have some updates over the source and the view on accumulative functions $ks$ and $kv$ respectively.\\
For $\cpginline{bx_1 \circ bx_2}$, there are only two $cpg$ calls. The first call $\cpginline{bx_1}$ requires parameter $(ks, id, s, construct\_dummy \ s)$ where $s$ and $ks$ are corresponding to the source and the update over source.
Because there is no real view here, we need to construct a dummy from the source. Then the continuation updating on this dummy should be initiated as an identity function. The first $cpg$ call is assigned to a 4-tuple $(ks_1, kv_1, s_1, v_1)$ where $s_1$ is redundant because this evaluated variable is not used in the later steps. In the next assignment, a 4-tuple $(ks_2, kv_2, s_2, v_2)$ is assigned by the second $cpg$ call which uses the input as $(kv_1, kv, v_1, v)$ where $kv_1$ and $v_1$ are obtained from the result of the first assignment, and $kv$ and $v$ come from the input. It is relatively similar to the second $pg$ call assignment in $\pginline{bx_1 \circ bx_2}$. After two $cpg$ calls, a function application, $ks_1 \ s_2$, is used to produce the updated source instead of calling recursively one more time like in $\pginline{bx_1 \circ bx_2}$.\\
Suppose that we have a source $s_0$ and a view $v_0$. The pair of the updated source and view $(s, v)$ where $s = \putbx{bx}{s_0}{v_0}$ and $v = \getbx{bx}{s_0}$ can be obtained using $cpg$ as follows:\\
    $\tab s \Leftarrow s_0; v \Leftarrow v_0; (ks, kv, s, v) \Leftarrow \cpg{bx}{\lambda \_.s}{id}{s}{v};\\
        \qtab (s; v)$\\
In general, the beginning of a continuation should be an identity function. However, to be able to use the function application to get the result of $\cpginline{bx_1 \circ bx_2}$, the accumulative function on the source $s$ needs to be initiated as a constant function $\lambda \_.s$.

Suppose the beginning of continuations $ks$ and $kv$ are $ks_0$ and $id$ respectively. Let's consider $\cpg{phead_1 \circ phead_2}{ks_0}{id}{s}{v}$ where $s = [[1,2,3], [\,], [4,5]]$ and $v = 100$. After the first two assignments in the definition of $cpg$ for the composition, we have: $ks_1 = f_2 \circ (con \ (fst \circ g_1 \circ id) \ (snd \circ f_1 \circ ks_0))$ and $s_2 = [100,2,3]$ where $f_1 = \lambda (s::ss).(s,ss)$, $f_2 = \lambda (s,ss).(s::ss)$, $g_1 = \lambda v.(v,())$. \\
Then: $ks_1 \ s_2 = (f_2 \circ (con \ (fst \circ g_1 \circ id) \ (snd \circ f_1 \circ ks_0))) \ s_2 \\
\tab = f_2 \ (\ (fst(g_1(id(s_2))) \ , \ snd(f_1(ks_0(s_2))))\ )\\
\tab = fst(g_1(id(s_2))) :: snd(f_1(ks_0(s_2))) = [100,2,3] :: snd(f_1(ks_0([100,2,3])))$\\
If $ks_0$ is an identity function, $ks_1 \ s_2 = [100,2,3] :: [2,3]$. This is an unexpected result when we target it to be the updated source. If $ks_0 = \lambda \_.s$ where $s = [[1,2,3], [\,], [4,5]]$, the result will be what we want to see.$ks_1 \ s_2 = [100,2,3] :: [[\,], [4,5]] = [[100,2,3], [\,], [4,5]]$. Through the above example, using the continuation $ks$ as a constant function at the beginning contributes to retain the discarded things in the source.

@@

The problem for $cpg$ lies in redundant computation during evaluating. For keeping away such computations, we introduce an extension named $kpg$. While $cpg$ evaluate values eagerly, $kpg$ does the opposite. Every values are evaluated lazily in a computation of $kpg$. The input of $\kpginline{bx}$ is expanded to a 6-tuple $(ks, kv, ks', kv', s, v)$ where $ks$ and $kv$ keep the modification information, $s$ and $v$ hold evaluated values, and $ks'$ and $kv'$ are used for lazy evaluation of actual values. The output of this function is also a 6-tuple $(ks, kv, ks', kv', s, v)$.\\
Suppose that we have a source $s_0$ and a view $v_0$. The pair of the updated source and view $(s, v)$ where $s = \putbx{bx}{s_0}{v_0}$ and $v = \getbx{bx}{s_0}$ can be obtained using $kpg$ as follows:\\
    $\tab s \Leftarrow s_0; v \Leftarrow v_0; (ks, kv, ks', kv', s, v) \Leftarrow \kpg{bx}{\lambda \_.s}{id}{id}{id}{s}{v};\\
        \qtab (ks' \ s; kv' \ v)$\\
The beginning of accumulative functions $ks'$ and $kv'$ are set as identity, while $ks$ and $kv$ are initiated as the same with the corresponding ones in $cpg$.

@@

$\kpg{Skip \ h}{ks}{kv}{ks'}{kv'}{s}{v} =\\
    \tab s \Leftarrow ks' \ s; \quad v \Leftarrow kv' \ v; \quad ks' \Leftarrow id; \quad kv' \Leftarrow id;\\
    \tab \text{if} \ h \ s = v \ \text{then} \ (ks, kv, ks', kv', s, v) \ \text{else fail}$

$\kpg{Replace}{ks}{kv}{ks'}{kv'}{s}{v} = (kv, ks, kv', ks', v, s)$

$\kpg{\product{bx_1}{bx_2}}{ks}{kv}{ks'}{kv'}{s}{v} =\\
    \tab s \Leftarrow ks' \ s; \quad v \Leftarrow kv' \ v; \quad ks' \Leftarrow id; \quad kv' \Leftarrow id;\\
    \tab (ks_1, kv_1, ks_1', kv_1', s_1, v_1) \Leftarrow \kpg{bx_1}{fst \circ ks}{fst \circ kv}{fst \circ ks'}{fst \circ kv'}{s}{v};\\
    \tab (ks_2, kv_2, ks_2', kv_2', s_2, v_2) \Leftarrow \kpg{bx_2}{snd \circ ks}{snd \circ kv}{snd \circ ks'}{snd \circ kv'}{s}{v};\\
    \qtab ( con \ ks_1 \ ks_2, con \ kv_1 \ kv_2, con \ (ks_1' \circ fst) \ (ks_2' \circ snd),\\
    \qtab con \ (kv_1' \circ fst) \ (kv_2' \circ snd), \\
    \qtab (s_1, s_2), (v_1,v_2))$

$\kpg{RearrS \ f_1 \ f_2 \ bx}{ks}{kv}{ks'}{kv'}{s}{v} =\\
    \tab (ks, kv, ks', kv', s, v) \Leftarrow \kpg{bx}{f_1 \circ ks}{kv}{f_1 \circ ks'}{kv'}{s}{v};\\
    \qtab (f_2 \circ ks, kv, f_2 \circ ks', kv', s, v)$

$\kpg{RearrV \ g_1 \ g_2 \ bx}{ks}{kv}{ks'}{kv'}{s}{v} =\\
    \tab (ks, kv, ks', kv', s, v) \Leftarrow \kpg{bx}{ks}{g_1 \circ kv}{ks'}{g_1 \circ kv'}{s}{v};\\
    \qtab (ks, g_2 \circ kv, ks', g_2 \circ kv', s, v)$

$\kpg{Case \ cond_{sv} \ cond_{s} \ bx_1 \ bx_2}{ks}{kv}{ks'}{kv'}{s}{v} =\\
    \tab s \Leftarrow ks' \ s; \quad v \Leftarrow kv' \ v; \quad ks' \Leftarrow id ; \quad kv' \Leftarrow id;\\
    \tab \text{if} \ cond_{sv} \ s \ v \&\& \ cond_{s} \ s\\
    \tab \text{then} \ (ks, kv, ks', kv', s', v') \Leftarrow \kpg{bx_1}{ks}{kv}{ks'}{kv'}{s}{v}\\
    \tab \text{else} \ (ks, kv, ks', kv', s', v') \Leftarrow \kpg{bx_2}{ks}{kv}{ks'}{kv'}{s}{v}\\
    \tab \text{fi} \ cond_s \ (ks' \ s') \ \&\& \ cond_{sv} \ s \ (kv' \ v'); \ \text{return} \ (ks, kv, ks', kv', s', v')$

>>>>

\section{Tupling and lazy updates: xpg}
\subsection{Tupling: pg}

Another solution for saving intermediate states is tupling. If $put$ and $get$ are evaluated simultaneously, it is potential to reduce the number of recomputed $get$s. The following function, $pg$, accepts the pair of a source and a view as the input to produce a new pair that contains the actual result of the corresponding minBiGUL program.

@@

\noindent Now, let's see how we construct $pg$ recursively.

    \noindent $\tabs{2} \pg{Skip \ h}{s}{v} \overset{1}{=} (\sif{h \ s = v}{s}{\text{fail}}, h \ s) \\
        \tab \qtab \overset{2}{=} \sif{h \ s = v}{(s, h \ s)}{\text{fail}} \\
        \tab \qtab \overset{3}{=} \sif{h \ s = v}{(s, v)}{\text{fail}}$

The first equality is simply based on the definitions of $pg$, $\putbxinline{Skip \ h}$ and $\getbxinline{Skip \ h}$. The second one tuples two results of $put$ and $get$ in the body of the if-expression. This is a trick since in some cases, the result of $pg$ may be fail although there is no fail when evaluating $\getbxinline{Skip \ h}$. The last equality is relatively oblivious.

    \noindent $\tabs{2} \pg{Replace}{s}{v} = (v, s)$\\
    $\tabs{2} \pg{\product{bx_1}{bx_2}}{(s_1,s_2)}{(v_1,v_2)}\\
        \tab \qtab \overset{1}{=} ((\putbx{bx_1}{s_1}{v_1}),(\putbx{bx_2}{s_2}{v_2}), (\getbx{bx_1}{s_1}), (\getbx{bx_2}{s_2})) \\
        \tab \qtab \overset{2}{=} (s_1, v_1) \Leftarrow \pg{bx_1}{s_1}{v_1}; \\
            \qtab \tabs{2} \ \, (s_2, v_2) \Leftarrow \pg{bx_2}{s_2}{v_2}; \\
            \tab \qtabs{2} ((s_1,s_2), (v_1,v_2))$\\
    $\tabs{2} \pg{\rearrs{f_1}{f_2}{bx}}{s}{v} \overset{1}{=} (f_2 \ (\putbx{bx}{(f_1 \ s)}{v}), \getbx{bx}{(f_1 \ s)}) \\
        \tab \qtab \overset{2}{=} (s, v) \Leftarrow \pg{bx}{f_1 \ s}{v};\\
            \tab \qtabs{2} (f_2 \ s, v)$\\
    $\tabs{2} \pg{\rearrv{g_1}{g_2}{bx}}{s}{v} \overset{1}{=} (\putbx{bx}{s}{(g_1 \ v)}, g_2 \ (\getbx{bx}{s})) \\
        \tab \qtab \overset{2}{=} (s, v) \Leftarrow \pg{bx}{s}{g_1 \ v}; \\
            \tab \qtabs{2} (s, g_2 \ v)$

Constructions of $pg$ for the replacement, the product and the source/view rearrangements are very clear when just paring $put$ and $get$ respectively, then doing basic changes.

    \noindent $\tabs{2} \pg{\casebx{cond_{sv}}{cond_{s}}{bx_1}{bx_2}}{s}{v} \\
    \tabs{3} \overset{1}{=} (\text{if} \ {cond_{sv} \ s \ v} \qtabs{3} \tab \text{if} \ {cond_{s} \ s} \\
    \tabs{2} \qtab \text{then} \ s' \Leftarrow \putbx{bx_1}{s}{v} \qtab \tab \, \text{then} \ v' \Leftarrow \getbx{bx_1}{s} \\
    \tabs{2} \qtab \text{else} \ s' \Leftarrow \putbx{bx_2}{s}{v} \qtab \tab \ \, \text{else} \ v' \Leftarrow \getbx{bx_2}{s}\\
    \tabs{2} \qtab \text{fi} \ cond_{s} \ s'; \text{ return } s' \qtab , \tab \ \ \text{fi} \ {cond_{sv} \ s \ v'}; \text{ return } v')\\
    \tabs{3} \overset{2}{=} \text{if} \ {cond_{sv} \ s \ v} \ \&\& \ {cond_{s} \ s}\\
    \tabs{2} \qtab \text{then} \ (s', v') \Leftarrow \pg{bx_1}{s}{v}\\
    \tabs{2} \qtab \text{else} \ (s', v') \Leftarrow \pg{bx_2}{s}{v}\\
    \tabs{2} \qtab \text{fi} \ cond_{s} \ s' \ \&\& \ cond_{sv} \ s \ v'; \text{ return } (s',v')$

@@

    \noindent $\tabs{2} \pg{bx_1 \circ bx_2}{s}{v} \\
    \tabs{3} \overset{1}{=} (\putbx{bx_1}{s}{(\putbx{bx_2}{(\getbx{bx_1}{s})}{v})}, \getbx{bx_2}{(\getbx{bx_1}{s})}) \\
    \tabs{3} \overset{2}{=} v_1 \Leftarrow \getbx{bx_1}{s}; \qtabs{3} \ \, \overset{3}{=} (s_1, v_1) \Leftarrow \pg{bx_1}{s}{dummy};\\
        \tab \qtab (s_2, v_2) \Leftarrow \pg{bx_2}{v_1}{v}; \qtabs{2} (s_2, v_2) \Leftarrow \pg{bx_2}{v_1}{v};\\
        \tab \qtab (s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2}; \qtabs{2} (s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2};\\
            \qtab \tabs{2} (s_3, v_2) \qtabs{6} (s_3, v_2)\\
    \tabs{3} \overset{4}{=} (s_1, v_1) \Leftarrow \pg{bx_1}{s}{construct\_dummy \ s}; \\
        \tab \qtab (s_2, v_2) \Leftarrow \pg{bx_2}{v_1}{v}; \\
        \tab \qtab (s_3, v_3) \Leftarrow \pg{bx_1}{s_1}{s_2}; \\
            \qtab \tabs{2} (s_3, v_2)$

The construction of $\pginline{bx_1 \circ bx_2}$ can be considered as the soul of the $pg$ function. The first two equalities comes from mentioned definitions and some basic transformations. The third one rewrites $v_1 \Leftarrow \getbx{bx_1}{s}$ into $(s_1, v_1) \Leftarrow \pg{bx_1}{s}{dummy}$. This is possible when we consider $\getbx{bx_1}{s}$ as the second element of $\pg{bx_1}{s}{dummy}$ where $dummy$ is a special value that makes the $put \, [\![bx_1]\!]$ valid. We can construct a $dummy$ from the source by \textsc{GetPut} law in some cases or require a valid view from programmers. Since there is no real view, these things are necessary. The last equality changes $dummy$ by an application $construct\_dummy \ s$, and also substitutes $s$ in $(s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2}$ by $s_1$ which equals to $\putbx{bx_1}{s}{(construct\_dummy \ s)}$. The replacement of $s$ and $s_1$ is possible because of the \textsc{PutPut} law. Then, $v_3$ in the result pair $(s_3,v_3)$ equals $dummy$. So we can realize that there is no loss information when computing a composition.

@@

    \noindent $\cpg{Skip \ h}{ks}{kv}{s}{v} = \text{ if } h \ s = v \text{ then } (ks, kv, s, v) \text{ else fail}$

    \noindent $\cpg{Replace}{ks}{kv}{s}{v} = (kv, ks, v, s)$

    \noindent $\cpg{\product{bx_1}{bx_2}}{ks}{kv}{s}{v} =\\
        \tab (ks_1, kv_1, s_1, v_1) \Leftarrow \cpg{bx_1}{fst \circ ks}{fst \circ kv}{fst \ s}{fst \ v};\\
        \tab (ks_2, kv_2, s_2, v_2) \Leftarrow \cpg{bx_2}{snd \circ ks}{snd \circ kv}{snd \ s}{snd \ v};\\
        \qtab (con \ ks_1 \ ks_2, con \ kv_1 \ kv_2, (s_1,s_2), (v_1,v_2))$

    \noindent $\cpg{RearrS \ f_1 \ f_2 \ bx}{ks}{kv}{s}{v} =\\
        \tab (ks, kv, s, v) \Leftarrow \cpg{bx}{f_1 \circ ks}{kv}{f_1 \ s}{v};\\
        \qtab (f_2 \circ ks, kv, s, v)$

    \noindent $\cpg{RearrV \ g_1 \ g_2 \ bx}{ks}{kv}{s}{v} =\\
        \tab (ks, kv, s, v) \Leftarrow \cpg{bx}{ks}{g_1 \circ kv}{s}{g_1 \ v};\\
        \qtab (ks, g_2 \circ kv, ks, g_2 \ v)$

    \noindent $\cpg{Case \ cond_{sv} \ cond_{s} \ bx_1 \ bx_2}{ks}{kv}{s}{v} =\\
        \tab \text{if} \ cond_{sv} \ s \ v \ \&\& \ cond_{s} \ s\\
        \tab \text{then} \ (ks, kv, s', v') \Leftarrow \cpg{bx_1}{ks}{kv}{s}{v}\\
        \tab \text{else} \ (ks, kv, s', v') \Leftarrow \cpg{bx_2}{ks}{kv}{s}{v}\\
        \tab \text{fi} \ cond_{s} \ s' \ \&\& \ cond_{sv} \ s \ v'; \ \text{return} \ (ks, kv, s', v')$

    \noindent $\cpg{bx_1 \circ bx_2}{ks}{kv}{s}{v} =\\
        \tab (ks_1, kv_1, s_1, v_1) \Leftarrow \cpg{bx_1}{ks}{id}{s}{construct\_dummy \ s};\\
        \tab (ks_2, kv_2, s_2, v_2) \Leftarrow \cpg{bx_2}{kv_1}{kv}{v_1}{v};\\
            \qtab (ks_1 \circ ks_2, kv_2,  ks_1 \ s_2, v_2)$
\end{definition}

Apart from the last construction, the others are quite similar to the corresponding ones of $pg$, but have some updates over the source and the view on accumulative functions $ks$ and $kv$ respectively.

For $\cpginline{bx_1 \circ bx_2}$, there are only two $cpg$ calls. The first call $\cpginline{bx_1}$ requires parameter $(ks, id, s, construct\_dummy \ s)$ where $s$ and $ks$ are corresponding to the source and the update over source. Since there is no real view here, we need to construct a dummy from the source. Then the continuation updating on this dummy should be initiated as an identity function. The first $cpg$ call is assigned to a 4-tuple $(ks_1, kv_1, s_1, v_1)$ where $s_1$ is redundant because this evaluated variable is not used in the later steps. In the next assignment, a 4-tuple $(ks_2, kv_2, s_2, v_2)$ is assigned by the second $cpg$ call which uses the input as $(kv_1, kv, v_1, v)$ where $kv_1$ and $v_1$ are obtained from the result of the first assignment, and $kv$ and $v$ come from the input. It is relatively similar to the second $pg$ call assignment in $\pginline{bx_1 \circ bx_2}$. After two $cpg$ calls, a function application, $ks_1 \ s_2$, is used to produce the updated source instead of calling recursively one more time like in $\pginline{bx_1 \circ bx_2}$.

Suppose that we have a source $s_0$ and a view $v_0$. The pair of the updated source and view $(s, v)$ where $s = \putbx{bx}{s_0}{v_0}$ and $v = \getbx{bx}{s_0}$ can be obtained using $cpg$ as follows:

    $\tab s \Leftarrow s_0; v \Leftarrow v_0; (ks, kv, s, v) \Leftarrow \cpg{bx}{\lambda \_.s}{id}{s}{v};\\
        \tab \qtab (s; v)$

In general, the beginning of a continuation should be an identity function. However, to be able to use the function application to get the result of $\cpginline{bx_1 \circ bx_2}$, the accumulative function on the source $s$ needs to be initiated as a constant function $\lambda \_.s$.

Suppose the beginning of continuations $ks$ and $kv$ are $ks_0$ and $id$ respectively. Let's consider $\cpg{phead_1 \circ phead_2}{ks_0}{id}{s}{v}$ where $s = [[1,2,3], [\,], [4,5]]$ and $v = 100$. After the first two assignments in the definition of $cpg$ for the composition, we have: $ks_1 = f_2 \circ (con \ (fst \circ g_1 \circ id) \ (snd \circ f_1 \circ ks_0))$ and $s_2 = [100,2,3]$ where $f_1 = \lambda (s::ss).(s,ss)$, $f_2 = \lambda (s,ss).(s::ss)$, $g_1 = \lambda v.(v,())$. Then:

$ks_1 \ s_2 = (f_2 \circ (con \ (fst \circ g_1 \circ id) \ (snd \circ f_1 \circ ks_0))) \ s_2 \\
\tab \quad = f_2 \ (\ (fst(g_1(id(s_2))) \ , \ snd(f_1(ks_0(s_2))))\ )\\
\tab \quad = fst(g_1(id(s_2))) :: snd(f_1(ks_0(s_2))) = [100,2,3] :: snd(f_1(ks_0([100,2,3])))$

If $ks_0$ is an identity function, $ks_1 \ s_2 = [100,2,3] :: [2,3]$. This is an unexpected result when we target it to be the updated source. If $ks_0 = \lambda \_.s$ where $s = [[1,2,3], [\,], [4,5]]$, the result will be what we want to see.$ks_1 \ s_2 = [100,2,3] :: [[\,], [4,5]] = [[100,2,3], [\,], [4,5]]$. Through the above example, using the continuation $ks$ as a constant function at the beginning contributes to retain the discarded things in the source.

@@

The problem for $cpg$ lies in redundant computation during evaluating. For keeping away such computations, we introduce an extension named $kpg$. While $cpg$ evaluate values eagerly, $kpg$ does the opposite. Every values are evaluated lazily in a computation of $kpg$. The input of $\kpginline{bx}$ is expanded to a 6-tuple $(ks, kv, ks', kv', s, v)$ where $ks$ and $kv$ keep the modification information, $s$ and $v$ hold evaluated values, and $ks'$ and $kv'$ are used for lazy evaluation of actual values. The output of this function is also a 6-tuple $(ks, kv, ks', kv', s, v)$.\\
Suppose that we have a source $s_0$ and a view $v_0$. The pair of the updated source and view $(s, v)$ where $s = \putbx{bx}{s_0}{v_0}$ and $v = \getbx{bx}{s_0}$ can be obtained using $kpg$ as follows:

    $\tab s \Leftarrow s_0; v \Leftarrow v_0; (ks, kv, ks', kv', s, v) \Leftarrow \kpg{bx}{\lambda \_.s}{id}{id}{id}{s}{v};\\
        \tab \qtab (ks' \ s; kv' \ v)$

The beginning of accumulative functions $ks'$ and $kv'$ are set as identity, while $ks$ and $kv$ are initiated as the same with the corresponding ones in $cpg$.

@@

    \noindent $\kpg{Skip \ h}{ks}{kv}{ks'}{kv'}{s}{v} =\\
        \tab s \Leftarrow ks' \ s; \quad v \Leftarrow kv' \ v; \quad ks' \Leftarrow id; \quad kv' \Leftarrow id;\\
        \tab \text{if} \ h \ s = v \ \text{then} \ (ks, kv, ks', kv', s, v) \ \text{else fail}$

    \noindent $\kpg{Replace}{ks}{kv}{ks'}{kv'}{s}{v} = (kv, ks, kv', ks', v, s)$

    \noindent $\kpg{\product{bx_1}{bx_2}}{ks}{kv}{ks'}{kv'}{s}{v} =\\
        \tab s \Leftarrow ks' \ s; \quad v \Leftarrow kv' \ v; \quad ks' \Leftarrow id; \quad kv' \Leftarrow id;\\
        \tab (ks_1, kv_1, ks_1', kv_1', s_1, v_1) \Leftarrow \kpg{bx_1}{fst \circ ks}{fst \circ kv}{fst \circ ks'}{fst \circ kv'}{s}{v};\\
        \tab (ks_2, kv_2, ks_2', kv_2', s_2, v_2) \Leftarrow \kpg{bx_2}{snd \circ ks}{snd \circ kv}{snd \circ ks'}{snd \circ kv'}{s}{v};\\
        \qtab ( con \ ks_1 \ ks_2, con \ kv_1 \ kv_2, con \ (ks_1' \circ fst) \ (ks_2' \circ snd),\\
        \qtab con \ (kv_1' \circ fst) \ (kv_2' \circ snd), \\
        \qtab (s_1, s_2), (v_1,v_2))$

    \noindent $\kpg{RearrS \ f_1 \ f_2 \ bx}{ks}{kv}{ks'}{kv'}{s}{v} =\\
        \tab (ks, kv, ks', kv', s, v) \Leftarrow \kpg{bx}{f_1 \circ ks}{kv}{f_1 \circ ks'}{kv'}{s}{v};\\
        \qtab (f_2 \circ ks, kv, f_2 \circ ks', kv', s, v)$

    \noindent $\kpg{RearrV \ g_1 \ g_2 \ bx}{ks}{kv}{ks'}{kv'}{s}{v} =\\
        \tab (ks, kv, ks', kv', s, v) \Leftarrow \kpg{bx}{ks}{g_1 \circ kv}{ks'}{g_1 \circ kv'}{s}{v};\\
        \qtab (ks, g_2 \circ kv, ks', g_2 \circ kv', s, v)$

    \noindent $\kpg{Case \ cond_{sv} \ cond_{s} \ bx_1 \ bx_2}{ks}{kv}{ks'}{kv'}{s}{v} =\\
        \tab s \Leftarrow ks' \ s; \quad v \Leftarrow kv' \ v; \quad ks' \Leftarrow id ; \quad kv' \Leftarrow id;\\
        \tab \text{if} \ cond_{sv} \ s \ v \ \&\& \ cond_{s} \ s\\
        \tab \text{then} \ (ks, kv, ks', kv', s', v') \Leftarrow \kpg{bx_1}{ks}{kv}{ks'}{kv'}{s}{v}\\
        \tab \text{else} \ (ks, kv, ks', kv', s', v') \Leftarrow \kpg{bx_2}{ks}{kv}{ks'}{kv'}{s}{v}\\
        \tab \text{fi} \ cond_s \ (ks' \ s') \ \&\& \ cond_{sv} \ s \ (kv' \ v'); \ \text{return} \ (ks, kv, ks', kv', s', v')$
